---
title: "Reaction Transport Modelling Course - Conceptual Diagrams"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=AwG9fYz3das&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=3)

and/or read the following sections of the book:

* Chapter 2.1. Conceptual model

(You can download each chapter of this book via the e-book University library for free!)

## Conceptual model diagrams

```{r State_variables, echo=FALSE}
question("When drawing a conceptual diagram we use the following to depict STATE VARIABLES", type = "learnr_radio",
answer("rectangles", correct=TRUE),
answer("ovals"),
answer("arrows"),
answer("dashed arrows"),
incorrect = "Rectangles is the correct answer" 
  )
```

```{r processes, echo=FALSE}
question("When drawing a conceptual diagram we use the following to depict PROCESSES:", type = "learnr_radio",
answer("rectangles"),
answer("ovals"),
answer("arrows", correct=TRUE),
answer("dashed arrows"),
incorrect = "Arrows is the correct answer" 
  )
```

```{r forcing_functions, echo=FALSE}
question("When drawing a conceptual diagram we use the following to depict FORCING FUNCTIONS:", type = "learnr_radio",
answer("rectangles"),
answer("ovals"),
answer("arrows"),
answer("dashed arrows", correct=TRUE),
incorrect = "dashed arrows is the correct answer" 
  )
```


```{r derived_variables, echo=FALSE}
question("When drawing a conceptual diagram we use the following to depict DERIVED VARIABLES:", type = "learnr_radio",
answer("rectangles"),
answer("ovals", correct=TRUE),
answer("arrows"),
answer("dashed arrows"),
incorrect = "Ovals is the correct answer" 
  )
```

## Ingredients of a conceptual model

```{r ingredients, echo=FALSE}
question("When creating a conceptual model we additionally need to specify", type = "learnr_checkbox",
 answer("The spatial domain", correct=TRUE),
 answer("The model parameters", correct=FALSE),
 answer("Temperature at which the processes occur", correct=FALSE),
 answer("The temporal domain", correct=TRUE),
 answer("The boundaries of the domain", correct=TRUE),
 answer("The concentrations of the reactants", correct=FALSE),
 answer("The boundary conditions for each state variable", correct=FALSE),
 answer("The initial conditions for each state variable", correct=FALSE),

incorrect = "For the conceptual model the state variables, spatial and temporal domain, including the boundaries are specified - the parameter values, initial conditions and boundary conditions are specified when you apply the model",
correct = "For the conceptual model the state variables, spatial and temporal domain, including the boundaries are specified - the parameter values, initial conditions and boundary conditions are specified when you apply the model")
```

## TRUE or FALSE statements

```{r statements, echo=FALSE}
question("Check each box that has a TRUE statement", type = "learnr_checkbox",
 answer("Transport is a process that transfers mass or energy", correct=TRUE),
 answer("All modelled state variables MUST have the same units", correct=FALSE),
 answer("If possible, it is better to use a common currency for all state variables", correct=TRUE),
 answer("For chemical species it is equally simple to use grams than moles", correct=FALSE),
 answer("The natural unit of a chemical species A is moles of A per kilogram or moles of A per m3", correct=TRUE)
  )
```

## Growth of a mussel

Assume you make a model that describes the growth of an individual organism (say a mussel), where you take into account the biomass increase of the organism, but also its energy reserves.

```{r Organism_Currency, echo=FALSE}
question("What would be a good currency for such a model ?", type = "learnr_radio",
 answer("C (carbon)", correct=TRUE),
 answer("N (nitrogen)"),
 answer("A mix of N and C"),
 incorrect = "Carbon is used both for biomass and energy production, so this is the best currency",
 correct = "Carbon is used both for biomass and energy production, so this is the best currency"
 )
```

## Algal growth

Microscopic algae convert dissolved inorganic C (DIC), dissolved inorganic nitrogen (DIN), and phosphorus (DIP) into their biomass.  

```{r Algae_Currency, echo=FALSE}
question("What would be a good currency for a model that describes the dependency of algae on dissolved inorganic N, P and C?", type = "learnr_radio",
 answer("C (carbon)", correct=FALSE),
 answer("N (nitrogen)", correct=FALSE),
 answer("P (phoshorus)", correct=FALSE),
 answer("a mix of C, N, P", correct=TRUE),
 incorrect = "It makes no sense to describe DIP in another currency than P, and similarly DIN and DIC should be represented in N and C respectively",
 correct = "Indeed, DIC, DIN and DIP can only be represented in C, N and P currencies respectively")
```



## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

More exercises on creating conceptual models can be found in the RTM-package:

```
require(RTM)
RTMexercise("conceptual")
```
