---
title: "Reaction Transport Modelling Course - transport fluxes"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=zAj4m6tGDhQ&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=12)

and/or read the following sections of the book:

* Chapter 3.4. Transport in a One-Dimensional Model

(You can download each chapter of this book via the e-book University library for free!)

## Transport fluxes

Consider the following expression for transport flux:

$$
J =  v \cdot S
$$

```{r adv, echo=FALSE}
question("Which of the following sentences is true?", type = "learnr_checkbox",
 answer(sprintf("This expression describes flux due to advective transport"), correct = TRUE),
 answer(sprintf("This expression describes flux due to dispersion")), 
 answer(sprintf("This expression describes flux due to diffusion")), 
 answer(sprintf("This expression describes flux due to bioturbation")), 

 answer(sprintf("v is the flow rate of the medium (e.g. water), in $m^3s^{-1}$")),
 answer(sprintf("v is the bulk velocity of the medium (e.g. water), in $m~s^{-1}$"), correct = TRUE),
 answer(sprintf("v is the dispersion or diffusion coefficient of the medium (e.g. water), in $m^2s^{-1}$")),
 answer(sprintf("v is the bulk volume of the medium (e.g. water), in $m^{3}$")),
 answer(sprintf("S is the concentration of the medium, in $mol~m^{-3}$"), correct=TRUE),
 answer(sprintf("S is the volume of the medium, in $m^3$")),
 answer(sprintf("S is the concentration gradient of the medium, in $mol~m^{-4}$")),

  incorrect=sprintf("The flux J is in mol $m^{-2}s^{-1}$, the advection rate $v$ is in m $s^{-1}$, the concentration $S$ is in mol $m^{-3}$"),
correct=sprintf("The flux J is in mol $m^{-2}s^{-1}$, the advection rate $v$ is in m $s^{-1}$, the concentration $S$ is in mol $m^{-3}$")  
)
```

Consider the following expression for transport flux:

$$
J =  -D \cdot \frac{\partial C}{\partial x}
$$

```{r difflux, echo=FALSE}
question("Which of the following sentences is true?", type = "learnr_checkbox",
 answer(sprintf("This expression describes flux due to advective transport")),
 answer(sprintf("This expression describes flux due to dispersion"), correct = TRUE), 
 answer(sprintf("This expression describes flux due to diffusion"), correct = TRUE), 
 answer(sprintf("This expression describes flux due to bioturbation"), correct = TRUE), 

 answer(sprintf("D is the flow rate of the medium (e.g. water), in $m^3s^{-1}$")),
 answer(sprintf("D is the bulk velocity of the medium (e.g. water), in $m~s^{-1}$")),
 answer(sprintf("D is the dispersion coefficient of the medium (e.g. water), in $m^2s^{-1}$"), correct = TRUE),
 answer(sprintf("D is the diffusion coefficient of the medium (e.g. water), in $m^2s^{-1}$"), correct = TRUE),
 answer(sprintf("D is the bulk volume of the medium (e.g. water), in $m^{3}$")),
 answer(sprintf("C is the concentration of the medium, in $mol~m^{-3}$"), correct=TRUE),
 answer(sprintf("C is the volume of the medium, in $m^3$")),
 answer(sprintf("C is the concentration gradient of the medium, in $mol~m^{-4}$")),
 answer(sprintf("$\\frac{\\partial C}{\\partial x}$ is the concentration gradient of the medium, in $mol~m^{-4}$"), correct=TRUE),

  incorrect=sprintf("The flux J is in mol $m^{-2}s^{-1}$, the advection rate $v$ is in m $s^{-1}$, the concentration $S$ is in mol $m^{-3}$"),
correct=sprintf("The flux J is in mol $m^{-2}s^{-1}$, the advection rate $v$ is in m $s^{-1}$, the concentration $S$ is in mol $m^{-3}$")  
)
```

## True or False

```{r trantruefalse, echo=FALSE}
question("Which of the following sentences is true", type = "learnr_checkbox",
 answer(sprintf("Molecular diffusion always decreases the concentration (hence the negative sign)")),
 answer(sprintf("In molecular diffusion transport goes against the concentration gradient (hence the negative sign) "), correct=TRUE),
 answer("Advective transport is due to a [directional] movement of particles, whereas diffusive transport is due to a [random] movement of particles.", correct = TRUE),  
 answer("Advective transport is due to a [random] movement of particles, whereas diffusive transport is due to a [directional] movement of particles."), 
 answer("In molecular diffusion, the diffusion coefficient can be written as a [product] of a characteristic distance that a particle travels between collisions with other particles and the characteristic velocity of the particle", correct = TRUE),
 answer("In molecular diffusion, the diffusion coefficient can be written as a [sum] of a characteristic distance that a particle travels between collisions with other particles and the characteristic velocity of the particle")
)
```

## diffusion

```{r diffusion, echo=FALSE}
question("In molecular diffusion, if the concentration is constant there is", type = "learnr_checkbox",
 answer("a constant transport of substance"),
 answer("no transport of substance"),
 answer("no net transport of substance", correct=TRUE)
  )
```

## bioturbation

```{r bioturbation, echo=FALSE}
question("Which of the following sentences is true?", type = "learnr_checkbox",
 answer("Bioturbation can be described as a diffusion-like process on time scales that are [much shorter than] intervals between translocation events of individual particles."),
 answer("Bioturbation can be described as a diffusion-like process on time scales that are [of the same order as[] the intervals between translocation events of individual particles."),
 answer("Bioturbation can be described as a diffusion-like process on time scales that are [much longer than] intervals between translocation events of individual particles.", correct=TRUE)
  )
```

## scales

```{r scaling, echo=FALSE}
question("What is the relationship between the typical values of the dispersion coefficient (Ddisp), the molecular diffusion coefficient (Dmol), and the bioturbation mixing coefficient (Dbio)?", type = "learnr_checkbox",
 answer("Ddisp >> Dbio >> Dmol"),
 answer("Ddisp << Dbio << Dmol"),
 answer("Ddisp << Dmol << Dbio"),
 answer("Ddisp >> Dmol >> Dbio", correct = TRUE)
  )
```

## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

