---
title: "Reaction Transport Modelling Course - Chemical Rate Laws"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=loHVpLNQBZE&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=5)

and/or read the following sections of the book:

* Chapter 2.3. Formulation of chemical reaction
* Chapter 8. Multiple time scales and Equilibrium Processes

(You can download each chapter of this book via the e-book University library for free!)

## TRUE or FALSE statements

```{r statements, echo=FALSE}
question("Check each box that has a TRUE statement", type = "learnr_checkbox",
 answer("Stoichiometric coefficients in an elementary chemical reaction directly determine the exponents in the expression for the reaction rate", correct=TRUE),
 answer("Stoichiometric coefficients in any chemical reaction directly determine the exponents in the expression for the reaction rate", correct=FALSE),
 answer("Monod kinetics can be derived from an enzymatic reaction", correct=TRUE) 
  )
```

## Irreversible elementary reaction

Consider the following irreversible chemical reaction:
$$
2A + B \rightarrow A_2B
$$

```{r elementary, echo=FALSE}
question("Which of the following expressions correctly describe the reaction rate", type = "learnr_checkbox",
 answer(sprintf("$R=k \\times [A]^2 \\times [B]$"), correct=TRUE),
 answer(sprintf("$R=k \\times [A] \\times [B]^2$"), correct=FALSE),
 answer(sprintf("$R=-k \\times [A_2B]$"), correct=FALSE),
 answer(sprintf("$R=k \\times [A] \\times [B]^{1/2}$"), correct=TRUE),
 answer(sprintf("$R=k \\times [A]^{1/2} \\times [B]$"), correct=FALSE),
 incorrect =  "there are two correct answers; answer 1 assumes that the rate R is expressed in mole of B per time; answer 4 assumes that R is expressed in mole of A per time - it is simplest to express the rate per mole of B",  
correct =  "there are two correct answers; answer 1 assumes that the rate R is expressed in mole of B per time; answer 4 assumes that R is expressed in mole of A per time - it is simplest to express the rate per mole of B")

```

## One reversible reaction: rates

Assume that the following reversible chemical reaction is the only process affecting ammonia ($NH_3$), ammonium ($NH_4^+$) and protons ($H^+$).
$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

* the rate constant for the forward reaction is called $k_f$
* the rate constant for the backward reaction is called $k_b$

```{r ammonia_eq, echo=FALSE}
question("What is the correct expression for the rate of change of ammonia?", type = "learnr_radio",
 answer(sprintf("$\\frac{d[NH_3]}{dt}=k_f \\times [NH_4] - k_b \\times [NH_3] \\times [H^+]$"), correct=TRUE),
 answer(sprintf("$\\frac{d[NH_3]}{dt}=k_f \\times [NH_4]$"), correct=FALSE),
 answer(sprintf("$\\frac{d[NH_3]}{dt}=k_f \\times [NH_4] - k_b \\times [NH_3]$"), correct=FALSE)
)
```

```{r proton_diffeq, echo=FALSE}
question("What is the correct expression for the rate of change of protons?", type = "learnr_radio",
 answer(sprintf("$\\frac{d[H^+]}{dt}=k_f \\times [NH_4]$"), correct=FALSE),
 answer(sprintf("$\\frac{d[H^+]}{dt}=k_f \\times [NH_4] - k_b \\times [NH_3]$"), correct=FALSE),
 answer(sprintf("$\\frac{d[H^+]}{dt}=k_f \\times [NH_4] - k_b \\times [NH_3] \\times [H^+]$"), correct=TRUE)
)
```

## One reversible reaction: units

$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

* the rate constant for the forward reaction is called $k_f$
* the rate constant for the backward reaction is called $k_b$

```{r unitsKf, echo=FALSE}
question(sprintf("What are the units of $k_f$?"), type = "learnr_radio",
 answer(sprintf("$(molN~ m^{-3})^{-1}~s^{-1}$"), correct=FALSE),
 answer(sprintf("$s^{-1}$"), correct=TRUE),
 answer(sprintf("$molN~m^{-3}~s^{-1}$"), correct=FALSE),
 incorrect = sprintf("The unit of the rate is $molN~ m^{-3}~s^{-1}$, and it is expressed as $R_f = k_f \\cdot [NH_4^+]$ - with $[NH_4^+]$ in $molN~ m^{-3}$, $k_f$ must be in units of $s^{-1}$ "),
   correct = sprintf("The unit of the rate is $molN~ m^{-3}~s^{-1}$, and it is expressed as $R_f = k_f \\cdot [NH_4^+]$ - with $[NH_4^+]$ in $molN~ m^{-3}$, $k_f$ must be in units of $s^{-1}$ ")
)
```

```{r unitsKb, echo=FALSE}
question(sprintf("What are the units of $k_b$?"), type = "learnr_radio",
 answer(sprintf("$(molN~ m^{-3})^{-1}~s^{-1}$"), correct=TRUE),
 answer(sprintf("$s^{-1}$"), correct=FALSE),
 answer(sprintf("$molN~m^{-3}~s^{-1}$"), correct=FALSE),
 incorrect = sprintf("The unit of the rate is $molN~ m^{-3}~s^{-1}$, and it is expressed as $R_b = k_b \\cdot [NH_3][H^+]$ - with $[NH_3]$ and $[H^+]$ in $molN~ m^{-3}$, $k_b$ must be in units of $(molN~ m^{-3})^{-1}~s^{-1}$ "),
   correct = sprintf("The unit of the rate is $molN~ m^{-3}~s^{-1}$, and it is expressed as $R_b = k_b \\cdot [NH_3][H^+]$ - with $[NH_3]$ and $[H^+]$ in $molN~ m^{-3}$, $k_b$ must be in units of $(molN~ m^{-3})^{-1}~s^{-1}$ ")
)
```

## One reversible reaction: lump sums

Based on the following equilibrium reversible reaction.
$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

```{r eqStatements, echo=FALSE}
question("Which of the following statements is true", type = "learnr_checkbox",
 answer(sprintf("$\\frac{d([H^+]+[NH_4^+])}{dt} =0$"), correct=TRUE),
 answer(sprintf("$\\frac{d([NH_3]+[NH_4^+]}{dt} =0$"), correct=TRUE),
 answer(sprintf("$\\frac{d([NH_3]-[NH_4^+]}{dt} =0$"), correct=FALSE),
 answer(sprintf("$\\frac{d([NH_3]+[H^+]}{dt} =0$"), correct=FALSE),
 answer(sprintf("$\\frac{d([NH_3]-[H^+]}{dt} =0$"), correct=TRUE),
 incorrect=sprintf("the sums of species whose combined derivatives are 0 are good candidates as lumpsums to formulate the equilibrium reactions : these lumpsums are:  excess$H^+$ = $[H^+]+[NH_4^+]$, $\\sum NH_x = [NH_3]+[NH_4^+]$ and $NH_3alkalinity$ = $[NH_3] -[H^+]$ and excess_charge = $[NH_4^+]-[H^+]$ - Note that you will need only two of those lumpsums to reformulate the problem"), 
  correct=sprintf("the sums of species whose combined derivatives are 0 are good candidates as lumpsums to formulate the equilibrium reactions : these lumpsums are:  excess$H^+$ = $[H^+]+[NH_4^+]$, $\\sum NH_x = [NH_3]+[NH_4^+]$ and $NH_3alkalinity$ = $[NH_3] -[H^+]$ and excess_charge = $[NH_4^+]-[H^+]$ - Note that you will need only two of those lumpsums to reformulate the problem")
 )
```

## One reversible reaction: equilibrium constant

Based on the following equilibrium reversible reaction.
$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

with forward rate constant called $k_f$ and the backward $k_b$.

Assume that 

* the equilibrium constant for this reaction is called $K_{N}$, with units of $mol~N~m^{-3}$.
* the concentration of total ammonia is defined as $\sum NH_x = [NH_3] + [NH_4^+]$.
* the equilibrium concentration of ammonia and ammonium is denoted as $[NH_3]^*$ and $[NH_4^+]^*$.

```{r KN, echo=FALSE}
question("Which of the following statements is true", type = "learnr_checkbox",
 answer(sprintf("$K_{N} = \\frac{k_f}{k_b}$"), correct=TRUE),
 answer(sprintf("$K_{N} = \\frac{k_b}{k_f}$"), correct=FALSE),

 incorrect=sprintf("if you find it difficult to remember this formula and do not want to derive it, think about the units of $K_N$ (mol $m^{-3}$), and the units of $k_b$ ($(molN~ m^{-3})^{-1}~s^-1$), and of $k_f$, ($s^{-1}$)). Units of both sides of the equality should be the same"), 
  correct=sprintf("if you find it difficult to remember this formula and do not want to derive it, think about the units of $K_N$ (mol $m^{-3}$), and the units of $k_b$ ($(molN~ m^{-3})^{-1}~s^-1$), and of $k_f$, ($s^{-1}$). Units of both sides of the equality should be the same")
 )
```

## One reversible reaction: equilibrium formulation

Based on the following equilibrium reversible reaction.
$$
NH_4^+(aq) \leftrightarrow NH_3(aq) + H^+
$$

where

* the equilibrium constant for this reaction is called $K_{N}$, with units of $mol~N~m^{-3}$.
* the concentration of total ammonia is defined as $\sum NH_x = [NH_3] + [NH_4^+]$.
* the equilibrium concentration of ammonia and ammonium is denoted as $[NH_3]^*$ and $[NH_4^{+}]^*$.

```{r eqStatements2, echo=FALSE}
question("Which of the following statements is true", type = "learnr_checkbox",
 answer(sprintf("$[NH_3]^* = \\frac{K_N}{K_N+[H^+]}$"), correct=FALSE),
 answer(sprintf("$[NH_3]^* = \\frac{K_N}{K_N+[H^+]} \\sum NH_x$"), correct=TRUE),
 answer(sprintf("$[NH_3]^* = \\frac{[H+]}{K_N+[H^+]}\\sum NH_x$"), correct=FALSE),
 answer(sprintf("$[NH_4^+]^* = \\frac{K_N}{K_N+[H^+]}$"), correct=FALSE),
 answer(sprintf("$[NH_4^+]^* = \\frac{K_N}{K_N+[H^+]}\\sum NH_x$"), correct=FALSE),
 answer(sprintf("$[NH_4^+]^* = \\frac{[H+]}{K_N+[H^+]}\\sum NH_x$"), correct=TRUE),
 incorrect=sprintf("if you find it difficult to remember which formula applies for which species, a menomnic we use is that the positively charged species has protons in the nominator"), 
  correct=sprintf("if you find it difficult to remember which formula applies for which species, a menomnic we use is that the positively charged species has protons in the nominator")
  
 )
```

## Equilibrium formulation: bjerrum plot

Bjerrum plots show how the concentration (or fraction) of the various ionic forms of a lumpsum change as a function of pH.

For ammonia we write, with $K_N = 5.380823e-10$ mol $kg^{-1}$:
```{r}
KN     <- 5.380823e-10
SumNHx <- 1  # to make it a fraction

pH <- seq(from=2, to=12, length.out=50)
H  <- 10^{-pH}
NH3 <- KN/(KN+H)
plot(pH, NH3, type = "l", ylab = "pNH3", main = "bjerrum")
```
 
## Two reversible reactions: equilibrium

You are given the following two equilibrium reactions:

$$
Reaction~1: \qquad H_2CO_3 \leftrightarrow HCO_3^- + H^+
$$


$$
Reaction~2: \qquad HCO_3^- \leftrightarrow CO_3^{2-} + H^+
$$

Equilibrium constants for the reactions 1 and 2 are $K_1$ and $K_2$, respectively. Assume that the concentration of the total dissolved inorganic carbon ($DIC = [H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$) is constant and equal to [DIC].


```{r eq2Reactions, echo=FALSE}
question("Which of the following equations correctly describe the relationships between carbonic acid, bicarbonate, carbonate and protons in equilibrium?", type = "learnr_checkbox",
 answer(sprintf("$[HCO_3^-] = \\frac{K_1 \\times [H^+]}{[H^+] \\times [H^+] + K_1 \\times [H^+] + K_1 \\times K_2} DIC$"), correct=TRUE),
 answer(sprintf("$DIC = \\frac{K_1 \\times [H^+]}{[H^+] \\times [H^+] + K_1 \\times [H^+] + K_1 \\times K_2} [HCO_3^-]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] = \\frac{K_2 \\times [H^+]}{[H^+] \\times [H^+] + K_1 \\times [H^+] + K_1 \\times K_2} DIC$"), correct=FALSE),
 answer(sprintf("$[CO_3^{2-}] = \\frac{[H^+] \\times [H^+]}{[H^+] \\times [H^+] + K_1 \\times [H^+] + K_1 \\times K_2} DIC$"), correct=FALSE),
 answer(sprintf("$[CO_3^{2-}] = \\frac{K_1 \\times K_2}{[H^+] \\times [H^+] + K_1 \\times [H^+] + K_1 \\times K_2} DIC$"), correct=TRUE),

 incorrect=sprintf("The equilibrium equations all depend on the DIC concentration, so the second option cannot be true. Also, the sum of $[H_2CO_3]$ (carbonic acid conc), $[HCO_3^-]$ (bicarbonate) and $[CO_3^{2-}]$ carbonate have to be = DIC (by definition), so the third equation cannot be true as the nominator is not part of the denominator; as for carbonate, the correct formula can be remembered as follows: carbonate is the most negatively charged ion, so it has in the nominator the uncharged combination (K1*K2) - the other nominators involving either one or two protons"), 
  correct=sprintf("The equilibrium equations all depend on the DIC concentration, so the second option cannot be true. Also, the sum of $[H_2CO_3]$ (carbonic acid conc), $[HCO_3^-]$ (bicarbonate) and $[CO_3^{2-}]$ carbonate have to be = DIC (by definition), so the third equation cannot be true as the nominator is not part of the denominator; as for the carbonate concentration, the correct formula can be remembered as follows: carbonate is the most negatively charged ion, so it has in the nominator the uncharged combination (K1*K2) - the other nominators involving either one or two protons") 
  )
```

## Two reversible reactions: lumpsums

You are given the following two equilibrium reactions:

$$
Reaction~1: \qquad H_2CO_3 \leftrightarrow HCO_3^- + H^+
$$


$$
Reaction~2: \qquad HCO_3^- \leftrightarrow CO_3^{2-} + H^+
$$

To make use of the fast reversible reactions, we need to find combinations of species that are not affected by the above reactions.

```{r eq2Lumpsum, echo=FALSE}
question("Which of the following lump sums would be suitable to reformulate a model that involves these equilbirum reactions?", type = "learnr_checkbox",
 answer(sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] - [H^+]$"), correct=TRUE),
 answer(sprintf("$[HCO_3^-] + [CO_3^{2-}] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] - [CO_3^{2-}] + [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] - 2[CO_3^{2-}] + [H^+]$"), correct=FALSE),
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$"), correct=TRUE),
 answer(sprintf("$2[H_2CO_3] + [HCO_3^-] + [H^+]$"), correct=TRUE),
  
 incorrect=sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] - [H^+]$ is unaffected by reaction 1 (production of $HCO_3^-$ =+1, production of $H^+$ = -1), and unaffected by reaction 2: (consumption of $HCO_3^-$ : -1, production of $H^+$ : -1, production of $CO_3^{2-}$ : +2).  This lump sum is called the 'carbonate alkalinity' or 'excess negative charge'. 
 The fifth lum sum is the 'total dissolved inorganic carbon'.
 The last lump sum ($2[H_2CO_3] + [HCO_3^-] + [H^+]$) is never used in models, but would also be suitable - it accounts for the excess protons.
   "), 
correct=sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] - [H^+]$ is unaffected by reaction 1 (production of $HCO_3^-$ =+1, production of $H^+$ = -1), and unaffected by reaction 2: (consumption of $HCO_3^-$ : -1, production of $H^+$ : -1, production of $CO_3^{2-}$ : +2).  This lump sum is called the 'carbonate alkalinity' or 'excess negative charge'.  
 The fifth lum sum is the 'total dissolved inorganic carbon'.
 The last lump sum ($2[H_2CO_3] + [HCO_3^-] + [H^+]$) is never used in models, but would also be suitable - it accounts for the excess protons.
   ")
)
```


## Combined reversible and irreversible reactions

You are given the following reactions:

$$
Reaction~1: \qquad H_2CO_3 \leftrightarrow HCO_3^- + H^+ \\
Reaction~2: \qquad HCO_3^- \leftrightarrow CO_3^{2-} + H^+\\
Reaction~3: \qquad NH_4^+ \leftrightarrow NH_3 + H^+\\
Reaction~4: \qquad HNO_3 \leftrightarrow NO_3^- + H^+\\
Reaction~5: \quad CH_2O(NH_3)_{NC} + O_2 \rightarrow H_2CO_3 + NC \cdot NH_3\\
Reaction~6: \quad NH_3 + 2O_2 \rightarrow NO_3^- + H^+ + H_2O
$$

* Assume that reactions 1–4 are fast and reversible, whereas reactions 5–6 correspond to a slow irreversible process (mediated by microbes). 
* Reaction 5, which corresponds to aerobic mineralization of organic matter, occurs at a rate *Raero*
* Reaction 6, which corresponds to nitrification, occurs at a rate *Rnit*.

```{r eqmultLumpsum, echo=FALSE}
question("Which of the following lump sums would be suitable to reformulate a model that involves the equilbrium reactions 1-4?", type = "learnr_checkbox",
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$"), correct=TRUE),
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}] + [CH_2O]$"), correct=FALSE),
 answer(sprintf("$[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}] - [CH_2O]$"), correct=FALSE),
 answer(sprintf("$[NH_4^+] + [NH_3]$"), correct=TRUE),
 answer(sprintf("$[NO_3^-] + [NH_3]$"), correct=FALSE),
 answer(sprintf("$[NO_3^-] + [HNO_3]$"), correct=TRUE),
 answer(sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] +[NH_3] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] +[NO_3^-] - [H^+]$"), correct=FALSE),
 answer(sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] +[NO_3^-] + [NH_3] - [H^+]$"), correct=TRUE),
 answer(sprintf("$[HCO_3^-] + 2 [CO_3^{2-}] -[HNO_3] + [NH_3] - [H^+]$"), correct=TRUE),
  
 incorrect=sprintf("The lump sums should only include species that are affected by the fast equilibrium processes, which rules out answer 2 and 3.  The lump sum $[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$ is the summed carbonate concentration that only features in the first 2 reactions (and is conserved by them), Similarly, $[NH_4^+] + [NH_3]$ and $[NO_3^-] + [HNO_3]$ are the summed ammonium and nitrate concentrations, who only feature and are conserved in the reactions 3 and 4. One more lump sum is needed to constrain the proton concentration. As protons feature in reactions 1, 2, 3, and 4 this lump sum should be conserved by all these equilibrium reactions. The lump sum $[HCO_3^-] + 2 [CO_3^{2-}] - [H^+]$ is not suitable, as it is not conserved in reactions 3 and 4; only the last 2 equations are suitable"),
  
correct=sprintf("The lump sums should only include species that are affected by the fast equilibrium processes, which rules out answer 2 and 3.  The lump sum $[H_2CO_3] + [HCO_3^-] + [CO_3^{2-}]$ is the summed carbonate concentration that only features in the first 2 reactions (and is conserved by them), Similarly, $[NH_4^+] + [NH_3]$ and $[NO_3^-] + [HNO_3]$ are the summed ammonium and nitrate concentrations, who only feature and are conserved in the reactions 3 and 4. One more lump sum is needed to constrain the proton concentration. As protons feature in reactions 1, 2, 3, and 4 this lump sum should be conserved by all these equilibrium reactions. The lump sum $[HCO_3^-] + 2 [CO_3^{2-}] - [H^+]$ is not suitable, as it is not conserved in reactions 3 and 4; only the last 2 equations are suitable")
)
```

## combined equilibrium/slow reactions: full model
 
You are given the following equilibrium reactions:

$$
Reaction~1: \qquad H_2CO_3 \leftrightarrow HCO_3^- + H^+ \\
Reaction~2: \qquad HCO_3^- \leftrightarrow CO_3^{2-} + H^+\\
Reaction~3: \qquad NH_4^+ \leftrightarrow NH_3 + H^+\\
Reaction~4: \qquad HNO_3 \leftrightarrow NO_3^- + H^+
$$
and slow reactions:

$$
Reaction~5: \qquad CH_2O(NH_3)_{NC} + O_2 \rightarrow H_2CO_3 + NC \cdot NH_3\\
Reaction~6: \qquad NH_3 + 2O_2 \rightarrow NO_3^- + H^+ + H_2O
$$

and the lumpsums:
$$
DIC = [H_2CO_3] + [HCO_3^-] + [CO3^{2-}]\\
NH_x = [NH_4^+] + [NH_3]\\
H_xNO_3 = [HNO_3] + [NO_3^-]\\
TA = [HCO_3^−] + 2 [CO_3^{2−}] + [NH_3] + [NO_3^-] − [H^+]
$$

```{r eqmultirrev, echo=FALSE}
question("How would you account for the aerobic respiration rate, Raero and the nitrification rate Rnit?", type = "learnr_checkbox",
 answer(sprintf("$\\frac{dDIC}{dt} = Raero$"), correct=TRUE),
 answer(sprintf("$\\frac{dH2CO3}{dt} = Raero$"), correct=FALSE),
 answer(sprintf("$\\frac{dCH_2O(NH_3)_{NC}}{dt} = -Raero$"), correct=TRUE),
 answer(sprintf("$\\frac{dNH_3}{dt} = NC \\times Raero$"), correct=FALSE),
 answer(sprintf("$\\frac{dNH_3}{dt} = NC \\times Raero - Rnit$"), correct=FALSE),
 answer(sprintf("$\\frac{dNH_x}{dt} = NC \\times Raero - Rnit$"), correct=TRUE),
 answer(sprintf("$\\frac{dH_xNO_3}{dt} = Rnit$"), correct=TRUE),
 answer(sprintf("$\\frac{dTA}{dt} = Raero + Rnit$"), correct=FALSE),
 answer(sprintf("$\\frac{dTA}{dt} = NC \\times Raero + Rnit$"), correct=FALSE),
 answer(sprintf("$\\frac{dTA}{dt} = NC \\times Raero - Rnit$"), correct=TRUE),
  
 incorrect=sprintf("The slow reaction should be written as a function of the lump sums, which rules out answer 2, 4 and 5.  Answer 1 and 6 is written as a function of DIC, $NH_x$, and $H_xNO_3$ and is correct. The alkalinity definition used here is affected by the aerobic respiration only by the production of NH3. The effect of nitrification on the TA (as defined here) is via the consumption of $NH_3$ (-1TA), the production of $NO_3^-$ (+1TA), and the production of protons -as the latter contribute negatively to TA, this gives a net consumption of TA of 1."),
  
correct=sprintf("The slow reaction should be written as a function of the lump sums, which rules out answer 2, 4 and 5.  Answer 1 and 6 is written as a function of DIC, $NH_x$, and $H_xNO_3$ and is correct. The alkalinity definition used here is affected by the aerobic respiration only by the production of NH3. The effect of nitrification on the TA (as defined here) is via the consumption of $NH_3$ (-1TA), the production of $NO_3^-$ (+1TA), and the production of protons -as the latter contribute negatively to TA, this gives a net consumption of TA of 1.")
)
```


## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

