---
title: "Reaction Transport Modelling Course - Chemical Rate Laws: Irreversible Reactions"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first. Only the first part (0:00-5:21) is relevant for this tutorial. TEST.

![](https://www.youtube.com/watch?v=loHVpLNQBZE&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=5)

and/or read the following sections of the book:

* Chapter 2.3. Formulation of chemical reaction

(You can download each chapter of this book via the e-book University library for free!)

## true/false statements

```{r statements, echo=FALSE}
question("Check each box that has a TRUE statement", type = "learnr_checkbox",
 answer("Stoichiometric coefficients in an elementary chemical reaction directly determine the exponents in the expression for the reaction rate", correct=TRUE),
 answer("Stoichiometric coefficients in any chemical reaction directly determine the exponents in the expression for the reaction rate", correct=FALSE),
 answer("Monod kinetics can be derived from an enzymatic reaction", correct=TRUE) 
  )
```

## reaction rate

Consider the following irreversible chemical reaction:
$$
2A + B \rightarrow A_2B
$$

```{r elementary, echo=FALSE}
question("Which of the following expressions correctly describe the reaction rate", type = "learnr_checkbox",
 answer(sprintf("$R=r \\times [A]^2 \\times [B]$"), correct=TRUE),
 answer(sprintf("$R=r \\times [A] \\times [B]^2$"), correct=FALSE),
 answer(sprintf("$R=-r \\times [A_2B]$"), correct=FALSE),
 answer(sprintf("$R=r \\times [A] \\times [B]^{1/2}$"), correct=TRUE),
 answer(sprintf("$R=r \\times [A]^{1/2} \\times [B]$"), correct=FALSE),
 incorrect =  "There are two correct answers; answer 1 assumes that the rate R is expressed in mole of B per time; answer 4 assumes that R is expressed in mole of A per time - it is simplest to express the rate in terms of mole of B",  
correct =  "There are two correct answers; answer 1 assumes that the rate R is expressed in mole of B per time; answer 4 assumes that R is expressed in mole of A per time - it is simplest to express the rate in terms of mole of B")
```

## units

You are given the following irreversible chemical reaction:
$$
2A + B \rightarrow A_2B
$$

where concentrations of A, B and $A_2B$ are expressed in mol m$^{-3}$, and the reaction rate is: 
$$
R=r \times [A]^2 \times [B]
$$

```{r rate, echo=FALSE}
question("What is the unit of R?", type = "learnr_checkbox",
 answer(sprintf("$mol~m^{-3}$"), correct=FALSE),
 answer(sprintf("$mol~m^{-3} s^{-1}$"), correct=TRUE),
 answer(sprintf("$mol^2~m^{-6} s^{-1}$"), correct=FALSE)
)
```


```{r ct, echo=FALSE}
question("What is the unit of rate constant r?", type = "learnr_checkbox",
 answer(sprintf("$mol~m^{-3}$"), correct=FALSE),
 answer(sprintf("$mol~m^{-3} s^{-1}$"), correct=FALSE),
 answer(sprintf("$(mol~m^{-3})^{-2} s^{-1}$"), correct=TRUE)
)
```

## mass balance equation

With the reaction
$$
2A + B \rightarrow A_2B
$$

```{r massbal, echo=FALSE}
question("toggle the correct mass balances", type = "learnr_checkbox",
 answer(sprintf("$\\frac{d[A]}{dt}=-R^2$"), correct=FALSE),
 answer(sprintf("$\\frac{d[A]}{dt}=-2 \\times R$"), correct=TRUE),
 answer(sprintf("$\\frac{d[B]}{dt}= R$"), correct=FALSE),
 answer(sprintf("$\\frac{d[B]}{dt}=-R$"), correct=TRUE),
 answer(sprintf("$\\frac{d[A_2B]}{dt}=R$"), correct=TRUE),
 answer(sprintf("$\\frac{d[A_2B]}{dt}=R^2$")),
 answer(sprintf("$\\frac{d[A_2B]}{dt}=2 \\times R$"))
)
```

## R-implementation

We repeat the equations from previous exercise here:

$$
\frac{d[A]}{dt}=-2 \times R; \qquad
\frac{d[B]}{dt}= - R; \qquad
\frac{d[A_2B]}{dt}= R \\
R = r \times [A]^2 \times [B]
$$

It is quite easy to implement this chemical model in R, as shown here.

```{r, message=FALSE}
require(deSolve)   # package with the solution methods

y.ini <- c(A=1, B=1, A2B=0)  # [mol/m3] initial concentrations of the state variables
parms <- c(r=0.001)          # [(molA/m3)^{-2}/s] rate constant

# derivative function
elementaryReaction <- function(t, y, p){  #t=time, y=state variables, p=parameters
  with (as.list(c(y,p)),{
    # Rate expressions
    R <- r * A^2 * B         # molB/m3/s
    # Mass balance equations
    dA.dt   <- -2*R
    dB.dt   <- -  R
    dA2B.dt <-    R
    # return vector of derivatives and ordinary variables
    list(c(dA.dt, dB.dt, dA2B.dt),  # same order as definition of y.ini
      sumA = A+2*A2B,    #mol A/m3
      sumB = B+A2B       #mol B/m3
      )
  })
}
```

The model is run twice, with different values of the parameter (x 10):
```{r}
# model solution
times <- seq(from=0, to=3600, length.out=100)  # seconds
out   <- ode(y=y.ini, times=times, parms=parms, func=elementaryReaction)

p2    <- c(r=0.01)  # [(molA/m3)^{-2}/s] rate constant 
out2  <- ode(y=y.ini, times=times, parms=p2, func=elementaryReaction)

plot(out, out2, lty=1)
```

TASK: Run the model for the rate that proceeds twice as slow as the default parameter. 


```{r run2, exercise=TRUE, exercise.cap = "Three model runs", exercise.lines = 10}
# model solution
times <- seq(from=0, to=3600, length.out=100)  # seconds
out   <- ode(y=y.ini, times=times, parms=parms, func=elementaryReaction)

p2    <- c(r=0.01)  # [(molA/m3)^{-2}/s] rate constant 
out2  <- ode(y=y.ini, times=times, parms=p2, func=elementaryReaction)

out3  <-     ### RUN WITH r=0.005
plot(out, out2, out3, lty=1)
```

## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

More extensive exercises on chemical models, more specific on elementary reactions, can be found in the RTM R-package:

```
require(RTM)
RTMexercise("ozone")
```
## References

R Core Team (2020). R: A language and environment for statistical computing. R
  Foundation for Statistical Computing, Vienna, Austria. URL
  https://www.R-project.org/.

Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential
  Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25.
  URL http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09
  