---
title: "Reaction Transport Modelling Course - reaction transport in porous media"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=zAj4m6tGDhQ&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=14)

and/or read the following sections of the book:

* Chapter 3.4.6. One-Dimensional Diffusion in Porous Media (Sediments)

(You can download each chapter of this book via the e-book University library for free!)

## Reaction-Transport in an Estuary

The following expression describes a reaction - transport model for organic C in an estuary of length L

$$
\frac{\partial C}{\partial t} =  -\frac{\partial}{\partial x}(v \cdot C)+\frac{\partial}{\partial x}(D \frac{ \partial C}{\partial x}) - r \cdot C
$$

```{r GTE, echo=FALSE}
question("Which is a good boundary condition for C at the upstream boundary (x=0)?", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed")),
 answer(sprintf("A flux boundary"), correct = TRUE),
 answer(sprintf("An imposed concentration boundary"), correct = TRUE)
)  
```


```{r GTE2, echo=FALSE}
question("Which is a good boundary condition for C at the downstream boundary (x=L)?", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed")),
 answer(sprintf("A flux boundary"), correct = TRUE),
 answer(sprintf("An imposed concentration boundary"), correct = TRUE)
)  
```

## Reaction-Transport in a River

The following expression describes a reaction - transport model for organic C in a river of length L

$$
\frac{\partial C}{\partial t} =  -\frac{\partial}{\partial x}(v \cdot C) - r \cdot C
$$

```{r GTE3, echo=FALSE}
question("Which is a good boundary condition for C at the upstream boundary (x=0)? '", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed")),
 answer(sprintf("A flux boundary"), correct = TRUE),
 answer(sprintf("A zero-gradient boundary condition")),
 answer(sprintf("An imposed concentration boundary"), correct = TRUE)
)  
```


```{r GTE4, echo=FALSE}
question("Which is a good boundary condition for C at the downstream boundary (x=L)?", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed"), correct = TRUE),
 answer(sprintf("A flux boundary")),
 answer(sprintf("An imposed concentration boundary"))
)  
```

## Early Diagenesis

Consider a two-component model in a porous media with porosity $\phi$. 

* Dissolved $SO_4^{2-}$ is a solute,  whose concentration is in mol S $m^{-3}$ of liquid.
* Particulate organic matter (OM) is a solid, whose concentration is in mol C $m^{-3}$ of solid. 

The two components are involved in the "sulphate reduction" reaction according to the stoichiometry:
$$
OM + 0.5 SO_4^{2-} \rightarrow CO_2 + S^{2-} +..
$$

```{r earlydia1, echo = FALSE}
question("Which is a good boundary condition for OM at the sediment-water interface (x=0)? ", type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed")),
 answer(sprintf("A flux boundary"), correct = TRUE),
 answer(sprintf("A zero-gradient boundary condition")),
 answer(sprintf("An imposed concentration boundary"))
)  
```

```{r earlydia2, echo = FALSE}
question(sprintf("Which is a good boundary condition for $SO_4^{2-}$ at the sediment-water interface (x=0)? "), type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed")),
 answer(sprintf("A flux boundary")),
 answer(sprintf("A zero-gradient boundary condition")),
 answer(sprintf("An imposed concentration boundary"), correct = TRUE)
)  
```

```{r earlydia3, echo = FALSE}
question(sprintf("Which is a good boundary condition for C deep down in the sediment (x=$\\infty$)? "), type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed")),
 answer(sprintf("A flux boundary")),
 answer(sprintf("A zero-gradient boundary condition"), correct = TRUE),
 answer(sprintf("An imposed concentration boundary"))
)  
```

```{r earlydia4, echo = FALSE}
question(sprintf("Which is a good boundary condition for $SO_4^{2-}$ deep down in the sediment (x=$\\infty$)? "), type = "learnr_checkbox",
 answer(sprintf("No boundary condition needed")),
 answer(sprintf("A flux boundary")),
 answer(sprintf("A zero-gradient boundary condition"), correct = TRUE),
 answer(sprintf("An imposed concentration boundary"), correct = TRUE),
  correct = ("the zero-gradient boundary is usually chosen for deep concentrations of solutes in diagenetic models, but sometimes concentrations are imposed, e.g. if there has been a historic buildup of solutes at depth"),
  incorrect = ("the zero-gradient boundary is usually chosen for deep concentrations of solutes in diagenetic models, but sometimes concentrations are imposed, e.g. if there has been a historic buildup of solutes at depth")
)  
```


## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

