---
title: "Reaction Transport Modelling Course - partitioning between phases"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
description: "learning tutorial in preparation for the exam"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=CPme0Qz3L7Y&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=8)

(there are no corresponding chapters in the book)

## Mineral dissolution - quiz

```{r statements, echo=FALSE}
question(sprintf("The rate constant describing the dissolution of a mineral in water is $k_d$ = 2 mol $m^{-2} yr^{-1}$, and the rate constant describing the precipitation of that mineral is $k_p$ = 5 m $yr^{-1}$. What is the equilibrium concentration of the dissolved mineral in water?"), type = "learnr_checkbox",
 answer(sprintf("0.4 mol $m^{-3}$"), correct=TRUE),
 answer(sprintf("2.5 mol $m^{-3}$")),
 answer(sprintf("1 mol $m^{-3}$")),
 answer(sprintf("10 mol $m^{-3}$")),
 answer(sprintf("400 $\\mu$ mol"), correct=TRUE),
 incorrect = sprintf("the answer is $k_d$/$k_p$ - tip:if you do not remember if this would be $k_p/k_d$ or  $k_d$/$k_p$, think about the units: concentrations are in mol $m^{-3}$, and this unit is obtained with  $k_d$/$k_p$, not with $k_p$/$k_d$"),
 correct = sprintf("the answer is $k_d$/$k_p$ - tip:if you do not remember if this would be $k_p/k_d$ or  $k_d$/$k_p$, think about the units: concentrations are in mol $m^{-3}$, and this unit is obtained with  $k_d$/$k_p$, not with $k_p$/$k_d$")   
  )
```


## Mineral dissolution - exercise

There are [N] spherical amorphous silica particles suspended in a well-mixed volume of water ([V] liters). Initially, the silicic acid concentration is [Cini] $\mu mol~L^{-1}$, and all particles have the same diameter (Dini) um. Assume the rate constant for silica dissolution of kd = 1 mol m-2 yr-1 and for silica precipitation of kp = 1.5 m yr-1. Silica density is 2200 kg m-3, molar weight of silica is 60 g mol-1.

The R-code chunk below introduces the values of the parameter you will need. 

```{r dSidt-setup}
N    <- 400
V    <- 1
Cini <- 100  # umol/L
Dini <- 90   # um
kd   <- 1    # mol/m2/yr
kp   <- 1.5  # m/yr
dens <- 2200 # kg/m3
mW   <- 60   # g/mol
```

What is the initial rate of increase in the silicic acid concentration (dC/dt) in the water with the silica particles? Give your value in $\mu mol~L^{-1} yr^{-1}$.
Use R to estimate it - you will need to do this in different steps. 

As a reminder: the change in Si concentration over time is given by:
$$
\frac{dSi}{dt} = -k_p \times \frac {A}{V} \times (Si - Si_{eq})
$$
where $Si_{eq}$ is the equilibrium Si concentration.

tip: beware to use consistent units. Also think about the meaning of $A$.

```{r dSidt, exercise=TRUE, exercise.cap = "Initial rate of increase of Si", exercise.lines = 4 }
# fill in the equation
dSi.dt <- 
```

```{r dSidt-hint}
# This shows the sequence of calculations - replace NA with the correct formula 
Ceq    <- NA                     # equilibrium concentration (umol/L)
A      <- NA                     # initial total active surface (m2)
A.V    <- NA                     # area density (/m)
dSi.dt <- -kp*A.V*(Cini - Ceq)   # the change in time umol/L/yr
```

```{r dSidt-solution}
Ceq    <- kd/kp*1000                # equilibrium concentration (umol/L)
A      <- 4*pi*(Dini/2*1e-6)^2 * N  # initial total active surface (m2)
A.V    <- A/(V*1e-3)                # area density (/m)
dSi.dt <- -kp*A.V*(Cini - Ceq)      # the change in time umol/L/yr
```


## Air-sea exchange - quiz

```{r statements2, echo=FALSE}
question(sprintf("A well-mixed anoxic water column is in contact with the atmosphere. What is the initial rate of increase in the $O_2$ concentration in the water? Piston velocity is 1 m $d^{-1}$, $O_2$ solubility in the water is 300 $\\mu mol$ $L^{-1}$, water column height is 50 cm."), type = "learnr_checkbox",
 answer(sprintf("300 $\\mu$ mol $L^{-1}d^{-1}$")),
 answer(sprintf("150 $\\mu$ mol $L^{-1}d^{-1}$")),
 answer(sprintf("600 $\\mu$ mol $L^{-1}d^{-1}$"), correct=TRUE),
 incorrect = sprintf("this is calculated as $kp/h*(O2eq - 0)$"),
 correct = sprintf("this is calculated as $kp/h*(O2eq - 0)$")
  )
```

## Air-sea exchange - exercise

You plan to install in your garden a pond with goldfish. By swimming in the pond, the fish will mix the water in the pond, and by respiration they will decrease the dissolved $O_2$ concentration in the pond. This decrease in dissolved $O_2$ will be compensated by air-water exchange. Assume that goldfish respiration is unaffected if the dissolved $O_2$ concentration does not decrease below [pSat] (80%) of the air saturated value. Additionally assume the piston velocity of [kp] m d-1, the area of the pond of 4 m2, the average height of the water column of [h] m, Henry’s constant [k0] for dissolved O2 of 770 L Atm mol-1, molar fraction of O2 in the atmosphere of 21%, atmospheric pressure of 1 Atm, and the average goldfish respiration rate of 400 $\mu$ mol fish$^{-1}$ $d^{-1}$.

```{r Fish-setup}
pSat  <- 0.9  # -
kp    <- 1    # m/d
h     <- 1    # m
K0    <- 770  # L atm/mol
P     <- 1    # atm
pO2   <- 0.21 # molar fraction
Resp  <- 400  # umol/fish/d  - I THINNK THIS IS ALREADY IN MMOL
Area  <- 4    # m2
```

Tip: at the minimal oxygen concentration, the aeration rate and respiration rate should be the same.

* Estimate the total air-water exchange (amount of oxygen added to the pond per day), at the minimal oxygen concentration (units mmol/day). 
* You know the respiration of one fish. The total respiration of all fishes should equal the total air-water exchange.

Use the R chunk below to add your answer (N is the number of fishes):

```{r Fish, exercise=TRUE}
N <-
```


```{r Fish-hint}
# This shows the sequence of calculations - replace NA with the correct formula 
O2sat   <-  NA              # saturated oxygen concentration, mmol/m3
airsea  <-  NA              # mmol/m2/d - airsea flux
airseaT <-  NA              # mmol/d - total O2 input per day
Rmmol   <-  NA              # fish respiration, mmol/fish/day
N       <-  NA              # number of fish    
```

```{r Fish-solution}
O2sat   <- pO2/K0*P *1e6                # umol/L = mmol/m3
airsea  <- kp*(O2sat - pSat*O2sat)      # mmol/m2/d - airsea flux
airseaT <- airsea*Area                  # mmol/d - total O2 input per day
Rmmol   <- Resp/1000                    # mmol/fish/day
N       <- airseaT/Rmmol                # number of fish    
```

## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

