---
title: "Reaction Transport Modelling Course - reaction transport in porous media"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=zAj4m6tGDhQ&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=13)

and/or read the following sections of the book:

* Chapter 3.4.6. One-Dimensional Diffusion in Porous Media (Sediments)

(You can download each chapter of this book via the e-book University library for free!)

## General Transport Equation

Consider the following expression for a reaction - transport model:

$$
\frac{\partial C}{\partial t} =  -\frac{\partial}{\partial x}(v \cdot C)+\frac{\partial}{\partial x}(D \frac{ \partial C}{\partial x}) + R
$$

```{r GTE, echo=FALSE}
question("In the simplified reaction-transport equation above, the first term on the left hand side describes: ", type = "learnr_checkbox",
 answer(sprintf("the substance concentration")),
 answer(sprintf("the substance concentration gradient")),
 answer(sprintf("the rate of change in the substance concentration"), correct = TRUE)
)  
```


```{r GTE2, echo=FALSE}
question("the first term on the right hand side describes: '", type = "learnr_checkbox",
 answer(sprintf("the substance concentration")),
 answer(sprintf("the substance concentration gradient")),
 answer(sprintf("the flux gradient due to dispersive transport")),
 answer(sprintf("the flux gradient due to advective transport"), correct = TRUE)
)  
```

```{r GTE3, echo=FALSE}
question("the second term on the right hand side describes: '", type = "learnr_checkbox",
 answer(sprintf("the substance concentration")),
 answer(sprintf("the substance concentration gradient")),
 answer(sprintf("the flux gradient due to dispersive transport"), correct = TRUE),
 answer(sprintf("the flux gradient due to advective transport"))
)  
```

## Porosity

```{r por, echo=FALSE}
question("In water-saturated porous media such as sediments, porosity is defined as", type = "learnr_checkbox",
 answer("volume of liquid divided by volume of solid"),
 answer("volume of liquid divided by bulk volume", correct = TRUE),
 answer("volume of solid divided by volume of liquid"),
 answer("volume of solid divided by bulk volume")
)  
```

## Solid volume fraction

```{r svf, echo=FALSE}
question("In water-saturated porous media such as sediments, solid volume fraction is defined as", type = "learnr_checkbox",
 answer("volume of liquid divided by volume of solid"),
 answer("volume of liquid divided by bulk volume"),
 answer("volume of solid divided by volume of liquid"),
 answer("volume of solid divided by bulk volume", correct = TRUE),
 answer(" one - porosity", correct = TRUE)  
)  
```

## Effective diffusion coefficient

```{r effdif, echo=FALSE}
question("Which if the sentences is true?", type = "learnr_checkbox",
 answer("In porous media, the effective diffusion coefficient of solutes is [smaller] than the molecular diffusion coefficient.", correct=TRUE),
 answer("In porous media, the effective diffusion coefficient of solutes is [larger] than the molecular diffusion coefficient."),
  incorrect = sprintf("the effective diffusion coefficient is defined as the molecular diffusion coefficient $\\times$ the porosity - and porosity is always < 1"),
  correct = sprintf("the effective diffusion coefficient is defined as the molecular diffusion coefficient $\\times$ the porosity - and porosity is always < 1")
)  
```

## Reaction-transport in porous media

```{r rteqpor, echo=FALSE}
question(sprintf("Which of the following equations correctly describes the reaction diffusion equation of a dissolved substance C in a water-saturated porous medium with porosity $\\phi$?"), type = "learnr_checkbox",
 answer(sprintf("$\\frac{\\partial \\phi C}{\\partial t} =  \\frac{\\partial}{\\partial x}(\\phi D \\frac{ \\partial C}{\\partial x}) + \\phi R$"), correct=TRUE),
 answer(sprintf("$\\frac{\\partial \\phi C}{\\partial t} =  \\frac{\\partial}{\\partial x}(D \\frac{ \\partial (\\phi C)}{\\partial x}) + \\phi R$")),
 answer(sprintf("$\\frac{\\partial C}{\\partial t} =  \\frac{\\partial}{\\partial x}(\\phi D \\frac{ \\partial C}{\\partial x}) + \\phi R$")),
 answer(sprintf("$\\frac{\\partial \\phi C}{\\partial t} =  \\frac{\\partial}{\\partial x}(D \\frac{ \\partial C}{\\partial x}) + R$")),
 incorrect=sprintf("The mass balance equation is derived for the bulk concentration, $\\phi C$ (on left hand side). However, the diffusion takes place in the liquid phase, hence the term $\\phi D \\frac{ \\partial C}{\\partial x}$ for the diffusion, where $D \\frac{ \\partial C}{\\partial x}$ is diffusion in the liquid phase, and $\\phi D \\frac{ \\partial C}{\\partial x}$ is diffusion in the bulk phase"),  
  correct=sprintf("The mass balance equation is derived for the bulk concentration, $\\phi C$ (on left hand side). However, the diffusion takes place in the liquid phase, hence the term $\\phi D \\frac{ \\partial C}{\\partial x}$ for the diffusion, where $D \\frac{ \\partial C}{\\partial x}$ is diffusion in the liquid phase, and $\\phi D \\frac{ \\partial C}{\\partial x}$ is diffusion in the bulk phase")
)  
```

## Reaction-transport in porous media-II

```{r rteqpor2, echo=FALSE}
question(sprintf("The form of the reaction-diffusion equation in porous media is usually written as: $\\frac{\\partial C}{\\partial t} = \\frac{1}{\\phi_x}  \\frac{\\partial}{\\partial x}(\\phi_x D \\frac{ \\partial C}{\\partial x}) + R$ - this assumes: "), type = "learnr_checkbox",
 answer("that porosity is constant in time but variable in space", correct=TRUE),
 answer("that porosity is constant in space AND in time"),
 answer("that porosity is variable in space AND in time")
)  
```

## True/False questions

```{r rteqpor3, echo=FALSE}
question(sprintf("When porosity is constant in time AND in space, then we don't need to know the porosity, and the following equation fully determines the diffusion model in saturated sediments: $\\frac{\\partial C}{\\partial t} = \\frac{\\partial}{\\partial x}(D \\frac{ \\partial C}{\\partial x})$ "), type = "learnr_checkbox",
 answer("this is true as porosity does not play a role then", correct=FALSE),
 answer("this is not true if we want to know the flux across the interface", correct=TRUE),
 incorrect = "if written like the above equation, the flux across the interface needs to be multiplied with porosity to have the correct units of mol / m2 total surface / time ",
  correct = "if written like the above equation, the flux across the interface needs to be multiplied with porosity to have the correct units of mol / m2 total surface / time "
  )  
```

## Porosity conversions
```{r porconv1, echo = FALSE}
question(sprintf("Select correct formulas for converting between concentrations per bulk volume ($C_b$), per volume of liquid ($C_L$), and per volume of solid ($C_S$). Porosity is denoted by $\\phi$."),
 answer(sprintf("$C_b = \\phi \\cdot C_L$"), correct=TRUE),
 answer(sprintf("$C_L = \\phi \\cdot C_b$")),
 answer(sprintf("$C_b = (1-\\phi) \\cdot C_L$")),
 answer(sprintf("$C_L = (1-\\phi) \\cdot C_b$")),
 answer(sprintf("$C_b = \\phi \\cdot C_S$")),
 answer(sprintf("$C_S = \\phi \\cdot C_b$")),
 answer(sprintf("$C_b = (1-\\phi) \\cdot C_S$"), correct=TRUE),
 answer(sprintf("$C_S = (1-\\phi) \\cdot C_b$")),
 incorrect = (sprintf("the units of $\\phi \\cdot C_L$ are $volume~of~Liquid~per~bulk \\times mol~per~ volume~Liquid$. This is equal to $mol~per~bulk$, the units of $C_b$")),
   correct = (sprintf("the units of $\\phi \\cdot C_L$ are $volume~of~Liquid~per~bulk \\times mol~per~ volume~Liquid$. This is equal to $mol~per~bulk$, the units of $C_b$ "))  
)
```

## Early diagenesis

Consider a two-component model in a porous media with porosity $\phi$. 

* Dissolved $SO_4^{2-}$ is a solute (e.g.,  whose concentration is in mol S $m^{-3}$ of liquid.
* Particulate organic matter (OM) is a solid whose concentration is in mol C $m^{-3}$ of solid. 

The two components are involved in the "sulphate reduction" reaction according to the stoichiometry:
$$
OM + 0.5 SO_4^{2-} \rightarrow CO_2 + S^{2-} +..
$$
The rate of sulphate reduction, $R_{sr}$ is expressed in mol C $m^{-3}$ solid $d^{-1}$.

```{r porconv2, echo = FALSE}
question(sprintf("How would you estimate the rate at which $SO_4^{2-}$ is removed?"),
 answer(sprintf("$R_{SO_4^{2-}} = R_{sr} \\times (1-\\phi) / \\phi$")),
 answer(sprintf("$R_{SO_4^{2-}} = 0.5 \\times R_{sr} \\times (1-\\phi) / \\phi$"), correct=TRUE),
 answer(sprintf("$R_{SO_4^{2-}} = 2 \\times R_{sr} \\times (1-\\phi) / \\phi$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = R_{sr} \\times \\phi/(1-\\phi)$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = 0.5 \\times R_{sr} \\times \\phi/(1-\\phi)$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = 2 \\times R_{sr} \\times \\phi/(1-\\phi)$"), correct=FALSE),
 answer(sprintf("$R_{SO_4^{2-}} = R_{sr}$")),
 answer(sprintf("$R_{SO_4^{2-}} = 0.5 \\times R_{sr}$")),
 answer(sprintf("$R_{SO_4^{2-}} = 2 \\times R_{sr}$")),
 incorrect = "you need to convert from mol/m3 solid/d to mol/m3 liquid/d; only 0.5 mol of SO4 is consumed per mol of C respired",
  correct = "you need to convert from mol/m3 solid/d to mol/m3 liquid/d; only 0.5 mol of SO4 is consumed per mol of C respired"
  )
```

## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

