---
title: "The coupled cycles of carbon, nitrogen, oxygen and sulfur in marine sediments"
subtitle: "Exercises Accompanying the Course Reaction Transport Modelling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky"
date: "March 2021"
output: 
  pdf_document:
    toc: no 
  header-includes: 
    \usepackage{float} 
    \floatplacement{figure}{H} 
---

```{r child = 'diagenesis_Q.Rmd'}
```

```{r child = 'diagenesis_A.Rmd'}
```

# References

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for
  Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
  
Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R:
  Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09

Soetaert, Karline and Meysman, Filip (2012). Reactive transport in aquatic ecosystems: Rapid
  model prototyping in the open source software R Environmental Modelling & Software, 32,
  49-60.
  
Soetaert K. (2009).  rootSolve: Nonlinear root finding, equilibrium and steady-state
  analysis of ordinary differential equations.  R-package version 1.6  
  