---
title: "Reaction Transport Modelling Course - transport processes"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=4ZwKZiobm0Y&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=10)

and/or read the following sections of the book:

* Chapter 3. Spatial Components and transport

(You can download each chapter of this book via the e-book University library for free!)

## General transport equation


$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{A\cdot J}}{\partial x}
$$

The differential equation above describes the variation of the concentration of a substance in time and space along 1 dimension. 


```{r GTE2, echo=FALSE}
question("Which of the following sentences is true?", type = "learnr_checkbox",
answer("We call this equation the one-dimensional reaction equation"),
answer("We call this equation the one-dimensional transport equation"),
answer("We call this equation the one-dimensional reaction-transport equation", correct=TRUE),
answer("The reaction term is the first on the right hand side", correct=TRUE),  
answer("The reaction term is the second on the right hand side"),  
answer("The transport term is the first on the right hand side"),  
answer("The transport term is the second on the right hand side", correct=TRUE)  
)
```

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{A\cdot J}}{\partial x}
$$


```{r GTE, echo=FALSE}
question("Which of the following sentences is true?", type = "learnr_checkbox",
 answer(sprintf("C is the flux of the substance")),
 answer(sprintf("C is the concentration of the substance"), correct = TRUE),
 answer(sprintf("t is time in s"), correct = TRUE),
 answer(sprintf("t is space in m"), correct = FALSE),
 answer(sprintf("x is time in s"), correct = FALSE),
 answer(sprintf("x is space in m"), correct = TRUE),
 answer(sprintf("R denotes the concentration in mmol/m3"), correct = FALSE),
 answer(sprintf("R denotes the net production rate in mmol/m3"), correct = FALSE),
 answer(sprintf("R denotes the net production rate in mmol/m2/s"), correct = FALSE),
 answer(sprintf("R denotes the net production rate in mmol/m3/s"), correct = TRUE),
 answer(sprintf("R denotes the flux in x-direction, in mmol/m2/s"), correct = FALSE),
 answer(sprintf("J denotes the concentration in mmol/m3"), correct = FALSE),
 answer(sprintf("J denotes the net production rate in mmol/m3"), correct = FALSE),
 answer(sprintf("J denotes the net production rate in mmol/m2/s"), correct = FALSE),
 answer(sprintf("J denotes the net production rate in mmol/m3/s"), correct = FALSE),
 answer(sprintf("J denotes the flux in x-direction, in mmol/m2/s"), correct = TRUE),
 answer(sprintf("A is the length, in m, through which the substance is transported"), correct = FALSE),
 answer(sprintf("A is the surface area, in m2, through which the substance is transported"), correct = TRUE),
 answer(sprintf("A is the volume, in m3, through which the substance is transported"), correct = FALSE)
)  
```  

## Fluxes

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{A\cdot J}}{\partial x}
$$

```{r transport, echo=FALSE}
question("Which of the following sentences is true?", type = "learnr_checkbox",
 answer(sprintf("The transport flux J describes the volume of substance transported per volume per time")),
 answer(sprintf("The transport flux J describes the amount of substance transported per unit area per time"), correct=TRUE),
 answer(sprintf("The transport flux J describes the amount of substance transported per unit volume per time")),
 answer(sprintf("The transport flux J describes the amount of substance transported per unit length per time")),
 answer(sprintf("The SI unit of the transport flux is  $s^{-1}$")),  
 answer(sprintf("The SI unit of the transport flux is mol $m^{-2}$ $s^{-1}$"), correct=TRUE),
 answer(sprintf("The SI unit of the transport flux is mol $m^{-3}$ $s^{-1}$")),
 answer(sprintf("The SI unit of the transport flux is mol $m^{-1}$ $s^{-1}$")),  
incorrect=sprintf("each of the terms in the equation has a SI unit of mol  $m^{-3}$ $s^{-1}$, as the SI unit of C is mol $m^{-3}$, and the SI unit of t is $s$ (second). The nominator in the second term on the right therefore has to have the unit of mol $m^{-2}$ $s^{-1}$, which is indeed the correct unit of flux"),
correct=sprintf("each of the terms in the equation has a SI unit of mol  $m^{-3}$ $s^{-1}$, as the SI unit of C is mol $m^{-3}$, and the SI unit of t is $s$ (second). The nominator in the second term on the right therefore has to have the unit of mol $m^{-2}$ $s^{-1}$, which is indeed the correct unit of flux")  
)
```

## Simplified equation

Starting from 

$$
\frac{\partial{C}}{\partial t} = R - \frac{1}{A}\cdot \frac{\partial{A\cdot J}}{\partial x}
$$

we rather use:

$$
\frac{\partial{C}}{\partial t} = R - \frac{\partial{J}}{\partial x}
$$
```{r transport2, echo=FALSE}
question("The above equation applies when", type = "learnr_checkbox",
 answer(sprintf("The transport flux J is zero")),
 answer(sprintf("The area A is zero")),
 answer(sprintf("The area A is constant"), correct=TRUE),
incorrect="mathematically, when A is constant, it can be taken out of the integral of the first formula, which gives rise to the second equation",
correct="mathematically, when A is constant, it can be taken out of the integral of the first formula, which gives rise to the second equation"  
)
```


## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

