---
title: "Reaction Transport Modelling Course - Ecological Rate Laws"
author: "Karline Soetaert and Lubos Polerecky"
date: "June-2021"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

## Tutorial

You are advised to watch the following video first 

![](https://www.youtube.com/watch?v=CUckCAwNj1s&list=PLx8PHcDdmF-uD1Pr07TU9SzlhlGpfrxqn&index=9)

and/or read the following sections of the book:

* Chapter 2.5. Basic Formulation of Ecological Interactions
* Chapter 2.6. Coupled Model Equations
* Chapter 2.7. Model Simplifications

(You can download each chapter of this book via the e-book University library for free!)

## Ecological interactions: algal N-uptake

```{r Nuptake, echo=FALSE}
question("Algal N-uptake is stimulated by nitrate (NO3) and inhibited by ammonia (NH3) concentrations. What would be a suitable expression for the rate of algal growth?", type = "learnr_checkbox",
 answer(sprintf("$r \\times \\frac{[NO3]}{[NO3]+K_{NO3}} \\times [ALGAE] + \\frac{K_{NH3}}{[NH3]+K_{NH3}} \\times [ALGAE]$")),
 answer(sprintf("$r \\times \\frac{[NO3]}{[NO3]+K_{NO3}} \\times \\frac{K_{NH3}}{[NH3]+K_{NH3}} \\times [ALGAE]$"), correct=TRUE),
 answer(sprintf("$r \\times \\frac{[NO3]}{[NO3]+K_{NO3}} \\times \\frac{K_{NH3}}{[NH3]+K_{NH3}}$")),
 answer(sprintf("$r \\times \\frac{[NO3]}{[NO3]+K_{NO3}} \\times [NH3]$")),
 answer(sprintf("$r \\times \\frac{[NO3]}{[NO3]+K_{NO3}} \\times (1 - \\frac{[NH3]}{[NH3]+K_{NH3}}) \\times [ALGAE])$"), correct=TRUE),
 answer(sprintf("$r \\times \\frac{K_{NO3}}{[NO3]+K_{NO3}} \\times [ALGAE]$"))
  )
```

## Ecological interactions: predation

```{r predation, echo=FALSE}
question("Foxes (F) predate on rabbits (R). What is a suitable expression for the predation rate?", type = "learnr_checkbox",
 answer(sprintf("$r \\times F \\times R$"), correct=TRUE),
 answer(sprintf("$r \\times F \\times R^2$"), correct=FALSE),
 answer(sprintf("$r \\times F \\times (1-\\frac{F}{R})$"), correct=FALSE),
 answer(sprintf("$r \\times F \\times \\frac{R}{R+ K_R}$"), correct=TRUE),
 answer(sprintf("$r \\times F \\times \\frac{R^2}{R^2+ K_R^2}$"), correct=TRUE),
 answer(sprintf("$r \\times F \\times (1-\\frac{F}{K}) \\times R$"), correct=FALSE),
 incorrect =  "there are several correct answers; answer 1 assumes a linear functional response (type I); answer 4 assumes a type II and answer 5 a type III functional response",  
correct =  "there are several correct answers; answer 1 assumes a linear functional response (type I); answer 4 assumes a type II and answer 5 a type III functional response")

```

## Simple differential equations


```{r diffeq1, echo=FALSE}
question(sprintf("The change in time of a quantity M (in moles) is described by the differential equation: $\\frac{dM}{dt} = -r \\times M$, where r > 0. Which of the following statements is true"), type = "learnr_checkbox",
 answer("this describes zero-order dynamics"),
 answer("this describes first-order dynamics", correct = TRUE),
 answer("this describes second-order dynamics"),
 answer("this is a mix of first-order and second-order dynamics"),
 answer(sprintf("the unit of the rate parameter $r$ is $day$")),
 answer(sprintf("the unit of the rate parameter $r$ is $/day$"), correct=TRUE),
 answer(sprintf("the unit of the rate parameter $r$ is $mol/day$")),
 answer(sprintf("the unit of the rate parameter $r$ is $day/mol$")),
 answer("M does not grow"),
 answer("M decreases linearly with time"),
 answer("M decreases exponentially with time", correct=TRUE),
 answer("M increases linearly with time"),
 answer("M increases exponentially with time"),
 answer("M has logistic growth")
)  
```

```{r diffeq2, echo=FALSE}
question(sprintf("The change in time of a quantity M (in moles) is described by the differential equation: $\\frac{dM}{dt} = -r$, where r > 0. Which of the following statements is true"), type = "learnr_checkbox",
 answer("this describes zero-order dynamics", correct = TRUE),
 answer("this describes first-order dynamics"),
 answer("this describes second-order dynamics"),
 answer("this is a mix of first-order and second-order dynamics"),
 answer(sprintf("the unit of the rate parameter $r$ is $day$")),
 answer(sprintf("the unit of the rate parameter $r$ is $/day$")),
 answer(sprintf("the unit of the rate parameter $r$ is $mol/day$"), correct=TRUE),
 answer(sprintf("the unit of the rate parameter $r$ is $day/mol$")),
 answer("M does not grow"),
 answer("M decreases linearly with time", correct=TRUE),
 answer("M decreases exponentially with time"),
 answer("M increases linearly with time"),
 answer("M increases exponentially with time"),
 answer("M has logistic growth")
)  
```

```{r diffeq3, echo=FALSE}
question(sprintf("The change in time of a quantity M (in moles) is described by the differential equation: $\\frac{dM}{dt} = r \\times M \\times (1-\\frac{M}{K})$, where r > 0. Which of the following statements is true"), type = "learnr_checkbox",
 answer("this describes zero-order dynamics"),
 answer("this describes first-order dynamics"),
 answer("this describes second-order dynamics"),
 answer("this is a mix of first-order and second-order dynamics", correct = TRUE),
 answer(sprintf("the unit of the rate parameter $r$ is $day$")),
 answer(sprintf("the unit of the rate parameter $r$ is $/day$"), correct=TRUE),
 answer(sprintf("the unit of the rate parameter $r$ is $mol/day$")),
 answer(sprintf("the unit of the rate parameter $r$ is $day/mol$")),
 answer("M does not grow"),
 answer("M decreases linearly with time"),
 answer("M decreases exponentially with time"),
 answer("M increases linearly with time"),
 answer("M increases exponentially with time"),
 answer("M has logistic growth", correct=TRUE)
)  
```

## Interactive differential equations

```{r, echo=FALSE}
radioButtons("type", label = "Differential equation:", choices = list("dM/dt = -r" = "a" , "dM/dt = r" = "b", "dM/dt = -r M" = "c", "dM/dt = r M"= "d", "dM/dt = r M (1-M/K)"="e"))
sliderInput("r", "Rate parameter:", min = 0, max = 1, value = 0.1)

plotOutput("eqType")
```

```{r, context="server"}
output$eqType <- renderPlot({
  r  <- 0.5
  r2 <- input$r 
  K  <- 1e6
  M0 <- 1
  times <- seq(0 , 10, length.out = 100)
 
  Type <- input$type
  if (Type == "a"){
    M <- M0 - r*times
    M2 <- M0 - r2*times
    Lab <- c("Zero-order differential equation dM/dt= -r", 
             "Linear decrease")
  } else if (Type == "b"){
    M  <- M0 + r*times
    M2 <- M0 + r2*times
    Lab <- c("Zero-order differential equation dM/dt= r", 
             "Linear increase")
  } else if (Type == "c") {

    M <- M0*exp(- r*times)
    M2 <- M0*exp(- r2*times)
    
    Lab <- c("First-order differential equation dM/dt= -r M", 
             "Exponential decrease")
  } else if (Type == "d") {
    M <- M0*exp(r*times)
    M2 <- M0*exp(r2*times)
    Lab <- c("First-order differential equation dM/dt= r M", 
             "Exponential increase")
  } else if (Type == "e")  {
    times <- seq(0 , 100, length.out = 100)

    A  <- (K -M0)/M0
    M  <- K/(1 +A*exp(-r*times))
    M2 <- K/(1 +A*exp(-r2*times))
    Lab <- c("Carrying capacity formulation dM/dt = r M(1-M/K)",
    "Logistic growth")
  }

  par(mfrow = c(1,1), las = 1, oma= c(0,0,2,0))
  matplot(times, cbind(M, M2), type="l", lwd=2, lty=1, main = Lab[2])
  legend (col = 1:2, lwd = 2, legend = c(r, r2), title = "r", x = "left")
  mtext(outer = TRUE, line = -1, Lab[1], cex = 2)
})
```

## Logistic model

A population is growing according to the logistic model characterised by the carrying capacity of $K=10^6$ ind $km^{-2}$.

```{r logistic1, echo=FALSE}
question(sprintf("How would you best characterize the growth of the population when the population density is about $10^5$ ind $km^{-2}$?"), type = "learnr_radio",
 answer("The population growth is approximately linear", correct=FALSE),
 answer("The population growth is approximately exponential", correct=TRUE),
 answer("The population stays relatively constant"), 
 answer("The population shrinks exponentially towards K")
)  
```

```{r logistic2, echo=FALSE}
question(sprintf("How would you best characterize the growth of the population when the population density is about $5~10^5$ ind $km^{-2}$?"), type = "learnr_radio",
 answer("The population growth is approximately linear", correct=TRUE),
 answer("The population growth is approximately exponential"),
 answer("The population stays relatively constant"), 
 answer("The population shrinks exponentially towards K")
)  
```


```{r logistic3, echo=FALSE}
question(sprintf("How would you best characterize the growth of the population when the population density is about $8.9~10^5$ ind $km^{-2}$?"), type = "learnr_radio",
 answer("The population growth is approximately linear"),
 answer("The population growth is approximately exponential"),
 answer("The population stays relatively constant", correct=TRUE), 
 answer("The population shrinks exponentially towards K")
)  
```

```{r logistic4, echo=FALSE}
question(sprintf("How would you best characterize the growth of the population when the population density is  $0$ ind $km^{-2}$?"), type = "learnr_radio",
 answer("The population growth is approximately linear"),
 answer("The population growth is approximately exponential"),
 answer("The population stays relatively constant", correct=TRUE), 
 answer("The population shrinks exponentially towards K")
)  
```

## logistic curve

Select a value of P0 and see how the population changes initially

```{r, echo=FALSE}
sliderInput("P0", "Population Size:", min = 0, max = 2e6, value = 5e5, step=1e5)
plotOutput("carcapPlot")
```

```{r, context="server"}
output$carcapPlot <- renderPlot({
  r  <- 1
  K  <- 1e6
  P0 <- input$P0
  A  <- (K -P0)/P0
  times <- seq(0 , 1, length.out = 100)
  times2 <- seq(0 , 10, length.out = 100)
  P  <- K/(1 +A*exp(-r*times))
  P2 <- K/(1 +A*exp(-r*times2))
  par(mfrow = c(1,2), las = 1, oma= c(0,0,2,0))
  plot(times, P, type="l", lwd=2, main="initial")
  plot(times2, P2, type="l", lwd=2, main="longterm")
  mtext(outer = TRUE, line = -1, paste("Logistic model, P0 = ", P0), cex = 2)
})
```


## Finally

```{r feedback, echo = FALSE}
question("give your feedback ", type = "learnr_text", answer(" ", correct=TRUE), correct="thank you", incorrect = "thank you")
```

