---
title: 'Interactive Applications in R'
subtitle: "Reader Accompanying the Course Reaction Transport Modeling in the Hydrosphere"
author: "Karline Soetaert and Lubos Polerecky"
date: "26-3-2021"
output:
  pdf_document: default
  html_document: default
fontsize: 10pt
vignette: >
  %\VignetteIndexEntry{Interactive Applications in R: accessory material}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{utf8}
abstract: Here we show how to make interactive (web-based) applications in R, using the shiny package. This allows to explore how the model behaves as a function of the input parameters. We use the ozone model as an example.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To write interactive applications in R ("webpages"), you need to install and load the "shiny" package (Chang et al., 2020).


```{r, message=FALSE}
require(shiny)
```

## Model implementation in R

The ozone model is a rather simple model that displays some nice features.

```{r, message = FALSE}
require(deSolve)
```

```{r}
state <- c(O = 0, NO = 1.3e8, NO2 = 5e11, O3 = 8e11)  # initial conditions

default.parms <- list(
  k3     = 1e-11,     # [/(mol/d)] 
  k2     = 1e10,      # [/d]
  k1a    = 1e-30,     # [/d] note: k1 = k1a + k1b*radiation
  k1b    = 1,         # [/(microEinst/m2/s)/d]
  sigma  = 1e11,      # [mol/d]           NO emission rate
  maxrad = 1200       # [microEinst/m2/s] maximal radiation
)

Ozone <- function(t, state, params) {
  with (as.list(c(state, params)), {

  radiation <- max(0, sin(t*2*pi))*maxrad    # radiation at time t (if time in days)

  # Rate expressions
  k1 <- k1a + k1b*radiation
  R1 <- k1*NO2
  R2 <- k2*O
  R3 <- k3*NO*O3
  
  # Mass balances [moles/day]
  dO   <-  R1 - R2
  dNO  <-  R1      - R3 + sigma
  dNO2 <- -R1      + R3
  dO3  <-       R2 - R3

  list(c(dO, dNO, dNO2, dO3), 
       radiation = radiation)
  })
}
```

for the web interface, we will create sliders, so that we can change the parameter values, and see the effect on model outcome. It is most instructive if we compare these altered model runs with the default model run. We therefore run the model first with the default parameters:

```{r}
outtimes <- seq(from = 0, to = 5, length.out = 300)  # run for 5 days
Default  <- ode(y=state, parms=default.parms, func=Ozone, times=outtimes, method="vode")
```


# The webpage part

The code for interactive applications consist of a user interface (UI) and a server.

## The user interface

We choose a main page with a side bar as the layout for this webpage (*pageWithSidebar*). 

* The header panel contains the title of the webpage
* The side bar contains 
   - sliders that can be tweaked to change model parameter values (*sliderInput*). Note that each slider has a name that will be accessed in the server function. Here we choose the name of the parameter as the slidername.
   - a check box that, when checked will plot the default run next to the current model run; its name is 'defaultRun' (*checkboxInput*).
   - a button that, when hit, will reset the default parameter values; its name is 'resetButton' (*actionButton*).
* The main panel contains the plot of the model run.

```{r}
UI.O3 <- shinyUI(pageWithSidebar(      # Define UI (user interfae)

  # Application title
  headerPanel("The ozone model"),

  sidebarPanel(
   sliderInput(inputId="k3",  label = "log(NO+O3 reaction rate)",   
     min = -13, max = -9, step = 0.1, value = log10(default.parms$k3)),
   sliderInput(inputId="k2",       label = "log(O+O2 reaction rate)",      
     min = 8, max = 12, step = 0.1, value = log10(default.parms$k2)),
   sliderInput(inputId="k1b",    label = "light-dependent NO2 rate", 
     min = 0, max = 5, step = 0.01, value = default.parms$k1b),
   sliderInput(inputId="sigma", label = "log(NO emission rate):", 
     min = 0, max = 12, step = 0.1, value = log10(default.parms$sigma)),
   sliderInput(inputId="maxrad",   label = "maximal radiation",   
     min = 0, max = 5000, value = default.parms$maxrad),
   
   actionButton (inputId="resetButton", label="Reset Parameters"),
    
   checkboxInput(inputId="defaultRun",  label=strong("Add default run"), value=TRUE),
   br()   # ends without ','
  ),

  mainPanel(
      plotOutput("PlotOzone"))
))
```

## The server

In the server we write the code that is executed when hitting the buttons/change the sliders in the UI.

Here,

* function "observeEvent" will only be triggered when an action has been executed. For instance, observeEvent(input$resetButton,... will be executed only when the reset button is hit.
* function "reactive" will only be executed when any of the sliders has been changed.
* function "renderPlot" will put a figure on the main panel. Note that the name "PlotOzone" is used in the server function.

```{r}
Server.O3 <- shinyServer(function(input, output, session) {

  # -------------------
  # the 'reset' button
  # -------------------
  observeEvent(input$resetButton, {
    updateNumericInput(session, "k3",     value = log10(default.parms$k3))
    updateNumericInput(session, "k2",     value = log10(default.parms$k2))
    updateNumericInput(session, "k1b",    value = default.parms$k1b)
    updateNumericInput(session, "sigma",  value = log10(default.parms$sigma))
    updateNumericInput(session, "maxrad", value = default.parms$maxrad)
  })

 # Get the model parameters, as defined in the UI 
  getparms <- reactive( {
    parms        <- default.parms 
    parms$k3     <- 10^input$k3
    parms$k2     <- 10^input$k2   
    parms$k1b    <- input$k1b     
    parms$sigma  <- 10^input$sigma  
    parms$maxrad <- input$maxrad
    parms
  })

  # -------------------
  # the 'Plot' tab
  # -------------------

  output$PlotOzone <- renderPlot({     # will be visible in the main panel

   parms <- getparms() # Model parameters, as defined in the UI
   out   <- ode(y=state, parms=parms, func=Ozone, times=outtimes, method="vode")

   if (input$defaultRun) {  # the check box is true
      plot (out, Default, lwd = 2, las = 1, lty = 1,
            cex.main = 1.5, cex.axis = 1.25, cex.lab = 1.25)  
      legend("bottomright", legend = c("current", "default"), 
             cex = 1.5, col = 1:2, lty = 1)
    } else  
      plot (out, lwd = 2, las = 1, lty = 1,
            cex.main = 1.5, cex.axis = 1.25, cex.lab = 1.25)  
   })                             # end ouput$plot

})     # end of the definition of shinyServer
```

## Run

to run this app, first run all the R-code in the console and then 
write the following:

shinyApp(ui = UI.O3, server = Server.O3)

You can leave the application by pressing <ESC>

```{r}

```

# References

Winston Chang, Joe Cheng, JJ Allaire, Yihui Xie and Jonathan McPherson (2020). shiny: Web
  Application Framework for R. R package version 1.4.0.2.
  https://CRAN.R-project.org/package=shiny
  
Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential
  Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25. URL
  http://www.jstatsoft.org/v33/i09/ DOI 10.18637/jss.v033.i09
  